---
title: "R Boot Camp - Billboard Demo"
author: "Chris Woolery"
output: html_notebook
---

```{r setup, include=FALSE}
# Copy the proper format to the YAML section above
# output: ioslides_presentation
# output: html_notebook
# output: powerpoint_presentation

knitr::opts_chunk$set(echo = FALSE)
```

## 1 - Libraries, Environment, Custom Functions and Parameter Defaults

```{r include=FALSE}
# Optional memory clear
rm(list=ls())
# Disable Scientific Notation in printing
options(scipen=999)
# Unload All Packages
lapply(names(sessionInfo()$otherPkgs), function(pkgs)
  detach(
    paste0('package:', pkgs),
    character.only = T,
    unload = T,
    force = T
  ))

QuietLoad <- function(library) {
  suppressWarnings(suppressPackageStartupMessages(
    library(library, character.only=TRUE)))
}

# Load libraries
# Load libraries
QuietLoad('tidyverse')
QuietLoad('readxl')
#QuietLoad('writexl')
QuietLoad('odbc')
QuietLoad('DBI')
QuietLoad('clipr')
QuietLoad('kableExtra')
QuietLoad('lubridate')
#QuietLoad('scales')
#QuietLoad('caret')
#QuietLoad('multidplyr')
#QuietLoad('sqldf')

### USEFUL UNIVERSAL FUNCTIONS

# A better data summary function; good for unknown data sets 
DataSummary <- function(Dataset) {
  Dataset[sapply(Dataset, is.character)] <- lapply(Dataset[sapply(Dataset, is.character)], as.factor)  
  StatSet <- data.frame(summary(Dataset), stringsAsFactors = FALSE) %>%
    rename(remove = 1,
           ColName = 2,
           ColStats = 3) %>%
    mutate(ColName = str_replace_all(ColName, "\\s", ""),
           ColStats = as.character(ColStats)) %>%
    dplyr::select(ColName, ColStats) %>%
    filter(!is.na(ColStats))
  SumSet <- data.frame(
    ColStats = Dataset[sapply(Dataset, is.numeric)] %>% 
    replace(is.na(.), 0) %>%
    sapply(sum),
    stringsAsFactors = FALSE) %>%
    mutate(ColName = row.names(.),
           ColStats = as.character(paste0("Sum : ", ColStats))) %>%
    dplyr::select(ColName, ColStats)
  ResultSet <-  bind_rows(StatSet, SumSet) %>%
    left_join(StatSet %>% 
                distinct(ColName) %>% 
                mutate(ColOrder = row_number()),
              by = c("ColName" = "ColName")) %>%
    arrange(ColOrder) %>%
    dplyr::select(ColOrder, ColName, ColStats)
  return(ResultSet)
}

# Semi-Universal Excel Date Converter
# Sub-Function of as.ExcelDate do not call directly
as.ExcelDate_Scalar <- function(ExcelDate_Scalar) {
  if (!is.na(ExcelDate_Scalar)) {
    if (str_detect(ExcelDate_Scalar, "\\/")) {
      as.Date(ExcelDate_Scalar, format = "%m/%d/%Y")
    } else {
      if (is.POSIXct(ExcelDate_Scalar)) {
        as.Date(ExcelDate_Scalar)
      } else {
        if (!is.na(as.numeric(ExcelDate_Scalar))) {
          as.Date(as.numeric(ExcelDate_Scalar), origin = "1899-12-30")
        } else {
          as.Date("2999-01-01")
        }
      }
    }
  } else {# Condition for NA inputs
    as.Date(NA)
  }
}

# Semi-Universal Excel Date Converter
# Callable function will process a scalar or vector and return the same
as.ExcelDate <- function(ExcelDate) {
  ExcelDateList <- lapply(ExcelDate, function(x) as.ExcelDate_Scalar(x))
  ExcelDateVector <- unlist(ExcelDateList, use.names = FALSE) %>% 
    as.Date(origin = "1970-01-01")
  return(ExcelDateVector)
}

# The not very efficient but one size fits most date converter
as.ETLDate <- function(DateVar) {
  DateVar <- ifelse(is.numeric(DateVar), as.character(DateVar), DateVar)
  tryCatch(RevDate <- case_when(
                        str_detect(datelikevector, "\\d{8}") 
                         & str_extract(datelikevector, "(?<=\\d{4}).{4}") <= "1231"
                          ~ as.Date(datelikevector, format = "%Y%m%d"),
                        str_detect(datelikevector, "\\d{8}") 
                         & str_extract(datelikevector, "(?<=\\d{4}).{4}") > "1231"
                          ~ as.Date(datelikevector, format = "%m%d%Y"),
                        str_detect(datelikevector, "\\d{6}")  
                         & str_extract(datelikevector, "..") <= "12"
                          ~ as.Date(datelikevector, format = "%m%d%y"),
                        str_detect(datelikevector, "\\d{6}")  
                        & str_extract(datelikevector, "..") > "12"
                        ~ as.Date(datelikevector, format = "%y%m%d"),
                        str_detect(datelikevector, "\\d\\/\\d\\d\\d\\d") 
                          ~ as.Date(datelikevector, format = "%m/%d/%Y"),
                        str_detect(datelikevector, "\\d\\-\\d\\d\\d\\d") 
                          ~ as.Date(datelikevector, format = "%m-%d-%Y"),
                        str_detect(datelikevector, "\\d\\d\\d\\d\\-\\d") 
                          ~ as.Date(datelikevector, format = "%Y-%m-%d"),
                        str_detect(datelikevector, "\\d\\d\\d\\d\\/\\d") 
                          ~ as.Date(datelikevector, format = "%Y/%m/%d"),
                        TRUE ~ as.Date(NA)),
           error = RevDate <- as.Date(NA))
  return(RevDate)
}

DeComma <- function(DollarString) {
  str_replace_all(DollarString, "\\$|\\,", "") %>%
    as.numeric()
  
}


RJETBlue = "#003365"

PrettyTable = function(TableObject, TableTitle) {
    TableObject %>%
      kable("html", escape = FALSE,
            caption = paste0('<p style="color:black; font-size:18px">',
            TableTitle,
            '</p>')) %>%
        kable_styling("striped",
                      bootstrap_options = c("hover", "condensed"),
                      full_width = T) %>%
        row_spec(0, color = "white", background = RJETBlue) 
}

PrettyModelTable = function(ModelObject, TableTitle) {
    summary(ModelObject)$coefficients %>%
      round(3) %>%
      data.frame() %>%
      rownames_to_column() %>%
      rename(Variable = 1, StdError = 3, TValue = 4, PValue = 5) %>%
      bind_rows(tibble(Variable = " R Squared",
                       Estimate = summary(ModelObject)$r.squared,
                       StdError = NA, TValue = NA, PValue = NA)) %>%
      arrange(Variable) %>%
      mutate(Significant = case_when(is.na(PValue) ~ "",
                                 PValue <= .05 & !Variable == "(Intercept)" ~ "Yes",
                                 TRUE ~"")) %>%
    kable("html", escape = FALSE,
          caption = paste0('<p style="color:black; font-size:18px">',
          TableTitle,
          '</p>')) %>%
      kable_styling("striped",
                    bootstrap_options = c("hover", "condensed"),
                    full_width = T) %>%
      row_spec(0, color = "white", background = RJETBlue) 
}

ModelQuartiles = function(BasicLM) {
  BasicLM = OriginalLM
  #rm(BasicLM, QuartileDF)
  QuartileDF = tibble(
  Actual = BasicLM$model[,1],
  Pred = BasicLM$fitted.values,
  ActRank = PercRank(BasicLM$model[,1])) %>%
  mutate(ActQuartile = 
          case_when(
            ActRank <= 0.25 ~ "1st Quartile",
            ActRank <= 0.50 ~ "2nd Quartile",
            ActRank <= 0.75 ~ "3rd Quartile",
            ActRank  > 0.75 ~ "4th Quartile",
            TRUE ~ "Error"),
         PredQuartile = 
          case_when(
            Pred <= quantile(BasicLM$model[,1], 0.25) ~ "1st Quartile",
            Pred <= quantile(BasicLM$model[,1], 0.50) ~ "2nd Quartile",
            Pred <= quantile(BasicLM$model[,1], 0.75) ~ "3rd Quartile",
            Pred  > quantile(BasicLM$model[,1], 0.75) ~ "4th Quartile",
            TRUE ~ "Error"),
         Success = if_else(ActQuartile == PredQuartile, 1, 0)) %>%
  group_by(ActQuartile) %>%
  summarize(Count = n(),
            Success = sum(Success),
            Accuracy = Success / Count,
            .groups = "drop")  
}



```
 
```{r include=FALSE}

GIT_CharTable = "https://raw.githubusercontent.com/RJETAnalytics/TrainingDemo/main/CharTable.RDATA"
download.file(GIT_CharTable,"CharTable.RDATA", method="curl")
CharTable <- readRDS("CharTable.RDATA")

# Read a Billboard table where each row is a song and its' rank in the top 100 for every week since the 1950's
demo_BillboardRank <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/billboard.csv')

# Read a table with song attributes from the Spotify database with an ID that allows us to link them to the Billboard rank table
demo_BillboardFeatures <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-09-14/audio_features.csv')

# Clean up date fields
demo_BillboardRank = demo_BillboardRank %>%
  mutate(week_id = as.Date(week_id, format = "%m/%d/%Y"))

# Clean up character fields for unwanted values
demo_BillboardFeatures = demo_BillboardFeatures %>%
    mutate_if(is.character, function(x) chartr(CharTable$Source, CharTable$Translated, x))

Billboard_DF = demo_BillboardRank %>%
  dplyr::select(-c(song, performer)) %>%
  inner_join(demo_BillboardFeatures, by = c("song_id" = "song_id")) %>%
  filter(!is.na(spotify_track_popularity))

```

One Hit Wonders versus Career Successes

Career Successes tend to have a variety of song types on their first album but stick with closer going forward then adapt over the years

Look at one hit wonders versus career successes and count them over time.

How similar are songs by all artists that chart in the TOP 40 during the year

Look at the career successes over time and look at the similarity of their songs and how it evolves over the years

Could use Spotify API and pull in catalogs of OHW vs CS

--

We will use this article https://en.wikipedia.org/wiki/List_of_one-hit_wonders_in_the_United_States which defines one hit wonders as "an act that has won a position on Billboard's national, pop, Top 40 just once." 

-Note about split on featuring

```{r}

# The performer text commonly has 'Artist A featuring Artist B' especially in the Hip Hop Genre
# which would create many 'false' One Hit Wonders; therefore we will strip out Artist A
# with this custom function
StripPerformer = function(RawText) {
  SplitText = "\\sFeaturing|\\sfeaturing|\\(Featuring|\\(featuring"
  StrLength = as.numeric(length(RawText))
  NewText = if_else(str_detect(RawText, SplitText),
                                str_sub(RawText, 1, str_locate(RawText, SplitText)[1:StrLength] -1),
                                RawText)
  return(NewText)
}

# Summarize the data frame and flag the one-hit wonders versus multiple hits
BillboardSummaryDF = Billboard_DF %>%
  filter(peak_position <= 40 & peak_position == week_position) %>%
  mutate(performer = StripPerformer(performer)) %>%
  group_by(performer, song) %>%
  summarize(peak_position = max(peak_position),
            weeks_on_chart = max(weeks_on_chart),
            week_id = min(week_id),
            danceability = mean(danceability), 
            energy = mean(energy),
            key = mean(key),
            mode = mean(mode),
            speechiness = mean(speechiness),
            acousticness = mean(acousticness), 
            instrumentalness = mean(instrumentalness), 
            liveness = mean(liveness), 
            valence = mean(valence),
            tempo = mean(tempo),
            .groups = "drop") %>%
  group_by(performer) %>%
  mutate(hit_count = n()) %>%
  arrange(hit_count, performer, week_id) %>%
  mutate(hit_number = row_number(),
         year_id = year(week_id),
         hit_type = if_else(hit_count == 1, "One Hit", "Multiple Hits")) %>%
  relocate(hit_count, hit_number, .before = performer) %>%
  relocate(year_id, .before = week_id)


```


```{r fig.width=10}

DisplayDF = BillboardSummaryDF %>%
  group_by(performer, hit_type) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(hit_type) %>%
  summarize(`Performer Count` = n(), .groups = "drop") %>%
  rename(`Type` = hit_type)

PrettyTable(DisplayDF, paste("One Hit Wonders vs Multiple Hit Performers - Since ", min(BillboardSummaryDF$year_id)))


```


```{r}

DataDF = BillboardSummaryDF
Euclidian = F
rm(p, NumericDF, ColNames, RowCount, Distances)
SelectCols = c("danceability", "energy", "key", "mode", "speechiness", "acousticness", "instrumentalness",
               "liveness", "valence", "tempo")

Minkowski = function(DataDF, Euclidian, SelectCols) {
  # Define the parameter p for the Minkowski distance where 1 = Euclidian distance
  # and 2 = Manhattan distance
  p = if_else(Euclidian, 2, 1)

  # Reduce data frame to numeric only fields
  NumericDF = DataDF %>%
    ungroup() %>%
    select(where(is.numeric))
  ColNames = colnames(NumericDF)
  RowCount = nrow(NumericDF)

  # Create an empty matrix to store the distances
  Distances <- matrix(nrow = nrow(NumericDF), ncol = nrow(NumericDF))
  
  # Calculate the distances between all pairs of rows
  i = 1
  j = 1
  for (i in 1:RowCount) {
    for (j in 1:RowCount) {
      xi <- NumericDF[i,ColNames]
      xj <- NumericDF[j,ColNames]
      Distances[i,j] <- (sum((abs(xi - xj))^p))^(1/p)
    }
  }
  return(Distances)
}

TestDF = tibble(
  name = c("Larry", "Moe", "Curly", "Shemp"),
  x1 = c(2,4,6, 50),
  x2 = c(1,2,3, 50)
)


x = Minkowski(BillboardSummaryDF, F, SelectCols)

Minkowski(TestDF, F, c("x1", "x2"))


```

```{r}

sqrt(4^2 + 2^2)

```


```{r fig.width=10}
DisplayDF = BillboardSummaryDF %>%
  group_by(year_id, hit_type) %>%
  summarize(n = n(), .groups = "drop") 

DisplayDF %>%
  ggplot(aes(year_id, n)) +
  geom_line(aes(color = hit_type))

```


 
