---
title: "R Boot Camp"
author: "Chris Woolery"
output:
  html_document:
    df_print: paged
---

This class is for people who are new to R but also:

- Have intermediate to expert skills in Excel
- Are familiar with basic to intermediate statistics
- Have basic to intermediate skills in some programming language

We will generally be following "R For Data Science" at: https://r4ds.had.co.nz/ which is a much better comprehensive guide than this class.

This class will skip some concepts and accelerate you through this book covering about 50% of what is there.  I advise going back and skimming through the things that do not look familiar as they will absolutely come up later in your R journey.


# DAY 4 
```{r include=FALSE}

# Optional memory clear
rm(list=ls())
# Disable Scientific Notation in printing
options(scipen=999)
# Unload All Packages
lapply(names(sessionInfo()$otherPkgs), function(pkgs)
  detach(
    paste0('package:', pkgs),
    character.only = T,
    unload = T,
    force = T
  ))

library(tidyverse)
library(lubridate)

# A better data summary function; good for unknown data sets 
DataSummary <- function(Dataset) {
  Dataset[sapply(Dataset, is.character)] <- lapply(Dataset[sapply(Dataset, is.character)], as.factor)  
  StatSet <- data.frame(summary(Dataset), stringsAsFactors = FALSE) %>%
    rename(remove = 1,
           ColName = 2,
           ColStats = 3) %>%
    mutate(ColName = str_replace_all(ColName, "\\s", ""),
           ColStats = as.character(ColStats)) %>%
    dplyr::select(ColName, ColStats) %>%
    filter(!is.na(ColStats))
  SumSet <- data.frame(
    ColStats = Dataset[sapply(Dataset, is.numeric)] %>% 
    replace(is.na(.), 0) %>%
    sapply(sum),
    stringsAsFactors = FALSE) %>%
    mutate(ColName = row.names(.),
           ColStats = as.character(paste0("Sum : ", ColStats))) %>%
    dplyr::select(ColName, ColStats)
  ResultSet <-  bind_rows(StatSet, SumSet) %>%
    left_join(StatSet %>% 
                distinct(ColName) %>% 
                mutate(ColOrder = row_number()),
              by = c("ColName" = "ColName")) %>%
    arrange(ColOrder) %>%
    dplyr::select(ColOrder, ColName, ColStats)
  return(ResultSet)
}

 # Create a distance function between two gps points
DistMiles <- function(LAT, LNG, LAT1, LNG1) {
           Euclidean <- 2 * 3858.8 *
             asin(sqrt(
               sin((LAT1 * pi/180 - LAT * pi/180)/2)^2 + 
                      cos(LAT * pi/180) * 
                      cos(LAT1 * pi/180) * 
                      sin((LNG1 * pi/180 - LNG * pi/180)/2)^2))
           return(Euclidean)
}

```

Let's load up some new publicly available data

```{r}

# Read a table where every row is an AirBnb Location in Barcelona Spain from Kaggle
# https://www.kaggle.com/datasets/zakariaeyoussefi/barcelona-airbnb-listings-inside-airbnb
 
# We could read the file and have R auto-assign data types based on the first 1000 and clean up the column names
# but we'll do that in advance to save time
# janitor::clean_names(NameOfDataset, case = "upper_camel") was used to get clean up the column names
#

Airbnb_ColNames = c("RowID", "PropertyId", "HostId", "HostIsSuperhost", "HostListingsCount", "Neighbourhood", "Zipcode", "Latitude", "Longitude", "PropertyType", "RoomType", "Accommodates", "Bathrooms", "Bedrooms", "Beds", "Amenities", "Price", "MinimumNights", "HasAvailability", "Availability30", "Availability60", "Availability90", "Availability365", "NumberOfReviewsLtm", "ReviewScoresRating")
Airbnb_ColTypes = c("nnnlnccnnccnnnnccnlnnnnnn")

demo_Airbnb_Barcelona <- readr::read_csv('https://raw.githubusercontent.com/RJETAnalytics/TrainingDemo/main/demo_Airbnb_Barcelona.csv',
                     col_types = Airbnb_ColTypes,
                     col_names = Airbnb_ColNames,
                     skip = 1)

#Clean up Amenities column and create 1 row per Amenities Data Frame
Airbnb_Amenities = demo_Airbnb_Barcelona %>%
  select(PropertyId, Amenities) %>%
  mutate(Amenities = str_remove_all(Amenities, "\\[|\\]")) %>%
  # separate_rows splits a field from 1 row into many based on a delimiter
  separate_rows(Amenities, sep = ", ") %>%
  # remove all \\W aka non-word characters
  mutate(Amenities = str_remove_all(Amenities, "\\W"),
         Amenities = paste0("Has", Amenities),
         HasAmenity = 1)

AGGAirbnb_Property_Amenities = Airbnb_Amenities %>%
  group_by(PropertyId) %>%
  summarize(AmenityCount = n(), .groups = "drop")

#Re-aggregate and rank to top 25 amenities by occurance
AGGAirbnb_Amenities = Airbnb_Amenities %>%
  group_by(Amenities) %>%
  summarize(AmenityCount = n(), .groups = "drop") %>%
  arrange(desc(AmenityCount)) %>%
  mutate(AmenityRank = row_number()) %>%
  filter(AmenityRank <= 25)

#Join back to Property ID by Amenity which by inner join gives us only top 25
#then pivot wider so that we have 1 rows per property with 1/0 for presence of the amenity 
Airbnb_Amenities = Airbnb_Amenities %>%
  inner_join(AGGAirbnb_Amenities, by = "Amenities") %>%
  # pivot wider makes 25 rows into 1 row with 25 new columns
  pivot_wider(id_cols = c(PropertyId),
              names_from = Amenities, values_from = HasAmenity, 
              values_fn = sum, values_fill = 0) %>%
  janitor::clean_names(case = "upper_camel") %>%
  # change from 1/0 to TRUE/FALCE logical
  mutate_at(vars(2:26), ~ as.logical(pmin(., 1))) %>%
  left_join(AGGAirbnb_Property_Amenities, by = "PropertyId") %>%
  mutate(AmenityCount = if_else(is.na(AmenityCount), 0, AmenityCount)) %>% 
  relocate(AmenityCount, .after = PropertyId)

#Re-join our top 25 amenity columns to our original table and drop the junky amenity column
AirbnbDF = demo_Airbnb_Barcelona %>%
  select(-Amenities) %>%
  inner_join(Airbnb_Amenities, by = c("PropertyId"))  

# Barcelona Central Plaza "Plaça de Catalunya" will be considered the City Center
CityCenter = tibble(LAT = 41.386553466894135, LONG = 2.170042714240173)

# Clean up data types and create defaults for NA values
# Character fields like Property Type and Neighbourhood that have finite values
# are converted to factors so that they can be used for prediction
# Add distance to city center
AirbnbDF = AirbnbDF %>%
  filter(!is.na(Neighbourhood)) %>%
  select(-c(Zipcode)) %>%
  mutate(HostIsSuperhost = if_else(is.na(HostIsSuperhost), FALSE, HostIsSuperhost),
         HostListingsCount = if_else(is.na(HostListingsCount), 0, HostListingsCount),
         Neighbourhood = factor(Neighbourhood),
         PropertyType = factor(PropertyType),
         RoomType = factor(RoomType),
         Bathrooms = if_else(is.na(Bathrooms), 0, Bathrooms) %>% as.numeric(),
         Bedrooms = if_else(is.na(Bedrooms), 0, Bedrooms) %>% as.numeric(),
         Beds = if_else(is.na(Beds), 0, Beds) %>% as.numeric(),
         Price = str_remove_all(Price, "\\$|\\,") %>% as.numeric(),
         ReviewScoresRating = if_else(is.na(ReviewScoresRating),
                                      median(ReviewScoresRating, na.rm = T),
                                      ReviewScoresRating),
         CityCenterDistance = DistMiles(Latitude, Longitude, CityCenter$LAT, CityCenter$LONG)) %>%
  relocate(CityCenterDistance, .after = Longitude)

# Defaulting to "Other"
# Some values are so sparse that they won't be useful for our modeling so we will
# replace anything with less than
MinAllowable = 10
# using fct_collapse or factor collapse
AirbnbDF <- AirbnbDF %>%
  mutate(PropertyType = fct_collapse(PropertyType ,
                                     Other = AirbnbDF %>% 
                                       count(PropertyType) %>%
                                       filter(n < MinAllowable) %>%
                                       pull(PropertyType)),
         Neighbourhood = fct_collapse(Neighbourhood ,
                                     Other = AirbnbDF %>% 
                                       count(Neighbourhood) %>%
                                       filter(n < MinAllowable) %>%
                                       pull(Neighbourhood)))

# Define the non-predictive columns for later exclusion
# ID columns aren't useful, zip code and lat/long are dups to neighborhood
# Availability is always true and our pricing prediction isn't dependent
# on it's availability for a snapshot in time (maybe scarcity is a driver but ignore for now)

```

AGENDA

1. Barcelona Airbnb Properties

    a. Join our rank and features data and filter to relevant columns
    
    b. Explore the data and visualize
    
    c. Build a linear model
    
    d. See if the linear model is predictive for price using test and train splits
    
    
2. Discuss Other Model Types

    a. AOV
    
    b. Classification versus Regression
    
    c. Logistic Regression
    
    d. Tree Models and other more complex models
    
    
3. Missed Concepts

    a. If ... Then and case ... when will be covered while manipulating the data in the prediction datasets


Explore the AirbnbDF
```{r}

summary(AirbnbDF)

```

Explore the thing we want to predict which is the price
```{r}
AirbnbDF %>%
  ggplot() +
    geom_histogram(aes(Price), bins = 20) +
    labs(title = "Price Distribution",
         subtitle = paste0("Price range ", min(AirbnbDF$Price), " - ", max(AirbnbDF$Price)))
```

Given the extreem outliers in high price, we might reset our goal to predict prices for the lower 95%
```{r}
AirbnbDF %>%
  filter(Price <= quantile(AirbnbDF$Price, 0.95)) %>%
  ggplot() +
    geom_histogram(aes(Price), bins = 20) +
    labs(title = "Price Distribution - Up to 95th Percentile",
         subtitle = paste0("Price range ", min(AirbnbDF$Price), 
                           " - ", max(AirbnbDF$Price[AirbnbDF$Price<=quantile(AirbnbDF$Price, 0.95)])))


```


```{r}

AirbnbDF %>%
  filter(Price <= quantile(AirbnbDF$Price, 0.95)) %>%
  ggplot(aes(CityCenterDistance, Price)) +
    geom_point(color = "grey") +
    geom_smooth(method = "lm", formula = 'y ~ x', se = F) +
    theme_classic() +
    labs(title = "Price vs Distance to City Center")

AirbnbDF %>%
  filter(Price <= quantile(AirbnbDF$Price, 0.95)) %>%
  ggplot(aes(AmenityCount, Price)) +
    geom_point(color = "grey") +
    geom_smooth(method = "lm", formula = 'y ~ x', se = F) +
    theme_classic() +
    labs(title = "Price vs Amenity Count")

AirbnbDF %>%
  filter(Price <= quantile(AirbnbDF$Price, 0.95)) %>%
  ggplot(aes(Accommodates, Price)) +
    geom_point(color = "grey") +
    geom_smooth(method = "lm", formula = 'y ~ x', se = F) +
    theme_classic() +
    labs(title = "Price vs Number of Guests (Accommodates)")

AirbnbDF %>%
  filter(Price <= quantile(AirbnbDF$Price, 0.95)) %>%
  filter(MinimumNights <= 180) %>%
  ggplot(aes(MinimumNights, Price)) +
    geom_point(color = "grey") +
    geom_smooth(method = "lm", formula = 'y ~ x', se = F) +
    theme_classic() +
    labs(title = "Price vs Minimum Nights")

```


```{r}

AirbnbDF_Clean = AirbnbDF %>%
  filter(Price <= quantile(AirbnbDF$Price, 0.95)) %>%
  filter(MinimumNights <= 180) %>%
  relocate(Price, .after = HostId) %>%
  select(-RowID)

Airbnb_LM = lm(Price ~ CityCenterDistance + AmenityCount + Accommodates,
               data = AirbnbDF_Clean)

summary(Airbnb_LM)

```

This formula tells us that:

Price = 29.01431
        + -9.02948 * City Center Distance 
        +  0.29246 * Amenity Count  
        + 17.17523 * Accommodates

This formula explains 43.89% of the variance in price
There are 18772 degrees of freedom + 4 prediction coefficients for 18776 observations
The range of variability (p value) for 3 predictor coefficients is small enough to consider them statistically significant.  This does NOT mean that the effective size is large (note that the change in price for adding 1 amenity is +0.29)


Let's explore the relationship between our target, predictors and the regression model we've built.

Now add the data from the model object

```{r fig.width=10}

AirbnbDF_Clean %>%
  mutate(Pred_Price = Airbnb_LM$fitted.values,
         AbsErr = abs(Pred_Price - Price)) %>%
  ggplot(aes(Pred_Price, Price)) +
    geom_point(aes(color = AbsErr)) +
#    geom_smooth(method = "lm", formula = "y ~ x", se = FALSE, size = 2, color = "pink") +
    scale_color_gradient(low = "grey", high = "red") +
    theme_classic() +
    labs(title = "Predicted Price versus Actual Price")

```

We can store the model object
```{r}

SaveFile = "C:\\Users\\Chris.Woolery\\OneDrive - Republic Airways\\Documents\\2024\\Airbnb_LM.RDATA"

saveRDS(Airbnb_LM, SaveFile)

x = readRDS(SaveFile)

```

We can plot the residuals (error between predicted and actual) ourselves to see what the bias in the variance is

```{r}

hist(Airbnb_LM$residuals, breaks = 30)

```

An R also has some nice default plots for most models

```{r fig.width=10}

plot(Airbnb_LM)

```

```{r}

library(caret)

set.seed(46268)
PartIndex = createDataPartition(AirbnbDF_Clean$Price, times = 1,
                                 p = 0.7, list = FALSE)

AirbnbDF_Clean = AirbnbDF_Clean %>%
  mutate(RowID = row_number(),
         SetName = as.factor(if_else(RowID %in% PartIndex, "Train", "Test"))) %>%
  dplyr::relocate(c(RowID, SetName), .before = everything()) %>%
  filter(complete.cases(.))

AirbnbDF_Clean %>%
  filter(SetName == "Train") %>%
  pull(Price) %>%
  summary()

```

```{r}

AirbnbDF_Clean %>%
  filter(SetName == "Test") %>%
  pull(Price) %>%
  summary()

```


```{r}

AirbnbDF_Train = AirbnbDF_Clean %>%
  filter(SetName == "Train")

AirbnbDF_Test = AirbnbDF_Clean %>%
  filter(SetName == "Test")

```

Let's introduce the concept of [cross validation](https://en.wikipedia.org/wiki/Cross-validation_%28statistics%29#/media/File:K-fold_cross_validation_EN.svg) for our training data


```{r}

set.seed(46268)
price.cv.folds <- createMultiFolds(AirbnbDF_Train$Price, k = 5, times = 4)

price.cv.cntrl <- trainControl(method = "repeatedcv", number = 5,
                         repeats = 4, index = price.cv.folds)

Airbnb_LM_Small <- train(Price ~ CityCenterDistance + AmenityCount + Accommodates
                   + HostIsSuperhost,
                           data = AirbnbDF_Train,
                           method="lm",
                           trControl = price.cv.cntrl)

summary(Airbnb_LM_Small$finalModel)

```

But we want to show an example of using a full featured model

```{r}

NonPredictiveColumns = c("RowID", "SetName", "PropertyId", "HostId", "ZipCode", 
                         "Latitude", "Longitude", "HasAvailability",
                         "Availability30", "Availability60", "Availability90",
                         "Availability365", 
                         "Neighbourhood", "RoomType")

Airbnb_LM <- train(Price ~ ., 
                           data = AirbnbDF_Train  %>%
                                    select(-any_of(NonPredictiveColumns)),
                           method="lm",
                           trControl = price.cv.cntrl)

summary(Airbnb_LM$finalModel)


```


```{r}

#Example of what predict() returns
x = predict(Airbnb_LM, newdata = AirbnbDF_Test)

#Add predictions to training and test data
ResultDF = bind_rows(AirbnbDF_Train %>%
                        mutate(Pred = predict(Airbnb_LM, AirbnbDF_Train),
                               Err = Pred - Price,
                               AbsErr = abs(Err)), 
                     AirbnbDF_Test  %>%
                        mutate(Pred = predict(Airbnb_LM, AirbnbDF_Test),
                               Err = Pred - Price,
                               AbsErr = abs(Err))
                     ) %>%
  relocate(c(Pred, Err, AbsErr), .after = Price)

```


```{r fig.width=10}

ResultDF %>%
  filter(SetName == "Test") %>%
  ggplot(aes(Pred, Price)) +
    geom_point(aes(color = AbsErr)) + 
    scale_color_gradient(low = "grey", high = "red") +
    theme_classic()

```


```{r}

Summary_ResultDF = ResultDF %>%
  group_by(SetName) %>%
  summarize(MeanPred = round(mean(Pred), 3),
            MeanPrice = round(mean(Price), 3),
            ME = round(mean(Err), 3),
            MAE = round(mean(AbsErr), 3),
            MAPE = round(MAE / mean(Price), 3),
            RMSE = round(sqrt(mean(Err^2)), 3),
            PctErrUnder25 = sum(if_else(AbsErr / Price <= 0.25, 1, 0)) / n(),
            .groups = "drop") 

Summary_ResultDF

```

```{r}

DisplayDF = ResultDF %>%
  mutate(Timeframe = Availability30 / 30,
         Category = case_when(Err > quantile(Err, 0.75) & 
                              Timeframe > mean(Timeframe)  ~ "Underpriced - High Availability",
                              Err > quantile(Err, 0.75) & 
                              Timeframe <= mean(Timeframe) ~ "Underpriced - Low Availability",
                              Err <= quantile(Err, 0.25) & 
                              Timeframe > mean(Timeframe) ~ "Overpriced - High Availability",
                              Err <= quantile(Err, 0.25) & 
                              Timeframe <= mean(Timeframe) ~ "Overpriced - Low Availability",      
                              Timeframe <= mean(Timeframe) ~ "Fair Price - Low Availability",      
                              Timeframe > mean(Timeframe) ~  "Fair Price - High Availability",
                              TRUE ~ "Other") %>% factor(),
         ) 

DisplayDF%>%
  group_by(Category) %>%
  summarize(n = n(), .groups = "drop") %>%
  mutate(Percentage = n / sum(n)) %>%
  ggplot(aes(reorder(Category, Percentage), Percentage)) +
  geom_col(aes(fill = Category)) +
  coord_flip() +
  theme(legend.position = "none") +
  labs(title = "Price Model vs Availability",
       x = "", y = "")

```

You could view and explore some of the Underpriced properties
```{r}

DisplayDF %>%
  filter(Err > 50) %>%
  arrange(desc(Err)) %>%
  View()

```


Now some publicly available data for Classification

```{r}
# Read a table where every row is an bank customer with some data and if they left the bank
# https://www.kaggle.com/datasets/saurabhbadole/bank-customer-churn-prediction-dataset
 
# We could read the file and have R auto-assign data types based on the first 1000 and clean up the column names
# but we'll do that in advance to save time
# No column name clean-up was needed as these were already in upper camel
#

BankChurn_ColTypes = c("nncnccnnnnnnnn")

demo_BankChurn <- readr::read_csv('https://raw.githubusercontent.com/RJETAnalytics/TrainingDemo/main/demo_BankChurn.csv', col_types = BankChurn_ColTypes)

# Some general data clean-up
BankChurnDF = demo_BankChurn %>%
  mutate(Geography = factor(Geography),
         Gender = factor(Gender),
         HasCrCard = as.logical(HasCrCard),
         IsActiveMember = as.logical(IsActiveMember),
         Result = if_else(Exited == 1, "Exited", "Retained") %>% factor()) %>%
  select(-c(Exited, Surname, -RowNumber)) %>%
  relocate(Result, .before = everything())

```
 
Explore the data
```{r}

summary(BankChurnDF)

```
Compare the Result to some of our predictor variables

First Numeric Variables with box plots
```{r fig.width=10}

DisplayDF = BankChurnDF %>%
  select(CustomerId, Result, CreditScore, Age, Tenure, Balance, NumOfProducts, EstimatedSalary) %>%
  pivot_longer(cols = -c(CustomerId, Result), names_to = "Predictor_Name", values_to = "Predictor_Value")

DisplayDF %>%
  ggplot(aes(Result, Predictor_Value)) +
    stat_boxplot(geom ='errorbar', size = 0.2) + 
    geom_boxplot() +
    facet_wrap(~Predictor_Name, ncol = 2, scales = "free") +
    coord_flip() 

```

Then Factors with column plots
```{r fig.width=10}

DisplayDF = BankChurnDF %>%
  # We'll plan on having a count for retained vs Exited
  # and convert logical to factor to make plots
  mutate(Retained = if_else(Result == "Retained", 1, 0),
         Exited = if_else(Result == "Exited", 1, 0),
         HasCrCard = factor(HasCrCard),
         IsActiveMember = factor(IsActiveMember)) %>%
  # Select just factors
  select(Retained, Exited, Geography, Gender, HasCrCard, IsActiveMember) %>%
  # Unpivot
  pivot_longer(cols = -c(Retained, Exited), names_to = "Predictor_Name", values_to = "Predictor_Value") %>%
  
  group_by(Predictor_Name, Predictor_Value) %>%
  summarise(Retained = sum(Retained),
            Exited = sum(Exited),
            .groups = "drop") %>%
  mutate(Exited_Rate = Exited / (Exited + Retained))

DisplayDF %>%
  ggplot(aes(Predictor_Value, Exited_Rate)) +
    geom_col() +
    facet_wrap(~Predictor_Name, ncol = 2, scales = "free") +
    coord_flip() 

```

```{r}

set.seed(46268)
PartIndex = createDataPartition(BankChurnDF$Result, times = 1,
                                 p = 0.7, list = FALSE)

BankChurnDF_Clean = BankChurnDF %>%
  mutate(RowID = row_number(),
         SetName = as.factor(if_else(RowID %in% PartIndex, "Train", "Test"))) %>%
  dplyr::relocate(c(RowID, SetName), .before = everything()) %>%
  filter(complete.cases(.))

BankChurnDF_Train = filter(BankChurnDF_Clean, SetName == "Train")
BankChurnDF_Test = filter(BankChurnDF_Clean, SetName == "Test")

```

Create a logistic regression model
https://en.wikipedia.org/wiki/Logistic_regression

```{r}

NonPredictiveColumns_BankChurn = c("RowID", "SetName", "RowNumber", "CustomerId")

BankChurn_LOG <- train(Result ~ .,
                           data = BankChurnDF_Train %>%
                                    select(-any_of(NonPredictiveColumns_BankChurn)),
                           method="glm")

summary(BankChurn_LOG$finalModel)

```
```{r}

confusionMatrix(BankChurn_LOG)

```

Other Models

Decision Trees
https://en.wikipedia.org/wiki/Decision_tree_learning
```{r}

BankChurn_RPART <- train(Result ~ .,
                           data = BankChurnDF_Train %>%
                                    select(-any_of(NonPredictiveColumns_BankChurn)),
                           method="rpart")

BankChurn_RPART$finalModel

```

```{r}

confusionMatrix(BankChurn_RPART)

```

```{r}
#Example of what predict() returns
x = predict(BankChurn_LOG, newdata = BankChurnDF_Test)

#Or we could just predict a single value if it was truely new data
NewData = tibble(
  CreditScore = 543,
  Geography = "France",
  Gender = "Female",
  Age = 56,
  Tenure = 3,
  Balance = 42000,
  NumOfProducts = 1,
  HasCrCard = FALSE,
  IsActiveMember = FALSE,
  EstimatedSalary = 192000
)

predict(BankChurn_LOG, NewData)

#Add predictions to training and test data
BankChurn_ResultDF = bind_rows(BankChurnDF_Train %>%
                                mutate(Pred = predict(BankChurn_RPART, BankChurnDF_Train)), 
                               BankChurnDF_Test %>%
                                mutate(Pred = predict(BankChurn_RPART, BankChurnDF_Test))) %>%
  relocate(c(Pred), .after = Result)

```

Create a confusion matrix from the TEST data
```{r}

confusionMatrix(filter(BankChurn_ResultDF, SetName == "Test")$Pred, 
                filter(BankChurn_ResultDF, SetName == "Test")$Result)

```

What could you use the predictions for if you are the bank pretending that the test data is new data that you don't really know the result for yet?



END
